---
title: "Final Project STAT 184"
author: "Siddharth Lal"
output: html_notebook
---

### Research Question 

How do seasonal changes and temperature variations impact sales in the retail sector, particularly focusing on different product categories?

Understanding how sales fluctuate throughout the year is crucial for retailers to optimize inventory, marketing strategies, and staffing. This analysis will help identify seasonal trends and patterns that can influence retail business decisions.

### Data Processing 

**Dataset Descriptions**

Retail Sales Dataset
This synthetic dataset simulates a dynamic retail environment, capturing essential attributes such as Transaction ID, Date, Customer ID, Gender, Age, Product Category, Quantity, Price per Unit, and Total Amount. The data is fictional but mirrors real-world retail scenarios, enabling a broad exploration of sales trends, demographic influences, and purchasing behaviors.

**Walmart Sales Dataset**
The Walmart sales dataset includes sales data from 45 stores in the United States, covering the period from February 5, 2010, to November 1, 2012. The data includes weekly sales volume, holiday status, temperature, fuel price, consumer price index (CPI), and unemployment rate.

Both datasets offer a comprehensive view of sales performance, allowing us to explore various factors that influence retail sales and customer behavior."


#### Data Intake 

# R code and a written explanation to import your **TWO** data sets.

```{r}
#load libraries
library(tidyverse)
library(knitr)
library(ggplot2)
library(dplyr)
library(lubridate)
```

```{r}
# Read the CSV files and display head and summary to ensure they have loaded properly
retail_sales_dataset <- read.csv("retail_sales_dataset.csv")
walmart_sales_data <- read.csv("Walmart_Sales_Data.csv")

# Display the first few rows of each dataset
str(retail_sales_dataset)
str(walmart_sales_data)

# Summarize each dataset
summary(retail_sales_dataset)
summary(walmart_sales_data)
```


#### Data Wrangling 
```{r}
# Ensure the Date columns are in Date format
retail_sales_dataset$Date <- as.Date(retail_sales_dataset$Date, format="%Y-%m-%d")

# Convert Walmart sales dataset Date column to "%Y-%m-%d" format
walmart_sales_data$Date <- as.Date(walmart_sales_data$Date, format="%d-%m-%Y")

# Display the structure of the datasets
str(retail_sales_dataset)
str(walmart_sales_data)

# Check the converted date formats
head(retail_sales_dataset$Date)
head(walmart_sales_data$Date)

retail_sales_dataset$Date <- as.Date(retail_sales_dataset$Date, format="%Y-%m-%d")
walmart_sales_data$Date <- as.Date(walmart_sales_data$Date, format="%d-%m-%Y")
retail_sales_dataset$Date <- as.Date(retail_sales_dataset$Date, format="%Y-%m-%d")

# Split the dataset by product category
categories <- unique(retail_sales_dataset$Product.Category)
category_datasets <- split(retail_sales_dataset, retail_sales_dataset$Product.Category)

# Function to aggregate sales data by month for a given category
aggregate_sales_by_month <- function(data) {
  data %>%
    mutate(Month = floor_date(Date, "month")) %>%
    group_by(Month, Product.Category) %>%
    summarise(Average_Sales = mean(Total.Amount))
}

# Aggregate sales data by month for each category
category_sales_monthly <- lapply(category_datasets, aggregate_sales_by_month)

# Combine the aggregated data into a single data frame
combined_sales_monthly <- bind_rows(category_sales_monthly)
```

```{r}
##DATA WRANGLING FOR THE CLOTHING AND TEMPERATURE GRAPH
retail_sales_dataset$Date <- as.Date(retail_sales_dataset$Date, format="%Y-%m-%d")
walmart_sales_data$Date <- as.Date(walmart_sales_data$Date, format="%d-%m-%Y")

# Calculate the average temperature for each month in the Walmart dataset
walmart_temp_monthly <- walmart_sales_data %>%
  mutate(Month = month(Date, label = TRUE)) %>%
  group_by(Month) %>%
  summarise(Average_Temperature = mean(Temperature, na.rm = TRUE))

walmart_temp_monthly$Month <- factor(walmart_temp_monthly$Month, levels = month.abb)

# Extract month from the retail sales dataset
retail_sales_with_temp <- retail_sales_dataset %>%
  mutate(Month = month(Date, label = TRUE)) %>%
  left_join(walmart_temp_monthly, by = "Month")

# Ensure Month is ordered
retail_sales_with_temp$Month <- factor(retail_sales_with_temp$Month, levels = month.abb)

# Filter the retail sales dataset for clothing sales
clothing_sales <- retail_sales_with_temp %>%
  filter(Product.Category == "Clothing")

# Aggregate clothing sales by month and calculate the average sales and average temperature
clothing_sales_monthly <- clothing_sales %>%
  group_by(Month) %>%
  summarise(Average_Sales = mean(Total.Amount), Average_Temperature = mean(Average_Temperature))
```






## Data Visualization

```{r}
# Aggregate sales data by month and calculate average sales per month for retail sales
retail_sales_monthly <- retail_sales_dataset %>%
  mutate(Month = floor_date(Date, "month")) %>%
  group_by(Month) %>%
  summarise(Average_Sales = mean(Total.Amount))

# Aggregate sales data by month and calculate average sales per month for Walmart sales
walmart_sales_monthly <- walmart_sales_data %>%
  mutate(Month = floor_date(Date, "month")) %>%
  group_by(Month) %>%
  summarise(Average_Sales = mean(Weekly_Sales))

# Create the ggplot2 graph for retail sales
plot_retail <- ggplot(retail_sales_monthly, aes(x = Month, y = Average_Sales)) +
  geom_line(color = 'blue') +
  scale_x_date(labels = date_format("%b"), breaks = date_breaks("1 month")) +
  scale_y_continuous(labels = dollar) +
  labs(title = "Average Sales by Month (Retail Sales)",
       subtitle = "The chart shows average sales per month",
       x = "Month",
       y = "Average Sales (in $)") +
  theme_minimal() +
  theme(plot.title = element_text(size = 20, face = "bold"),
        plot.subtitle = element_text(size = 14),
        axis.text.x = element_text(angle = 45, hjust = 1))

# Create the ggplot2 graph for Walmart sales
plot_walmart <- ggplot(walmart_sales_monthly, aes(x = Month, y = Average_Sales)) +
  geom_line(color = 'red') +
  scale_x_date(labels = date_format("%b"), breaks = date_breaks("1 month")) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Average Sales by Month (Walmart Sales)",
       subtitle = "The chart shows average sales per month",
       x = "Month",
       y = "Average Sales (in $)") +
  theme_minimal() +
  theme(plot.title = element_text(size = 20, face = "bold"),
        plot.subtitle = element_text(size = 14),
        axis.text.x = element_text(angle = 45, hjust = 1))

# Display the plots
print(plot_retail)
print(plot_walmart)
```

```{r}
# Create the ggplot2 graph for each category
plot_category_sales <- ggplot(combined_sales_monthly, aes(x = Month, y = Average_Sales, color = Product.Category)) +
  geom_line() +
  scale_x_date(labels = date_format("%b"), breaks = date_breaks("1 month")) +
  scale_y_continuous(labels = dollar) +
  labs(title = "Average Sales by Month for Each Product Category",
       subtitle = "The chart shows average sales per month for each product category",
       x = "Month",
       y = "Average Sales (in $)",
       color = "Product Category") +
  theme_minimal() +
  theme(plot.title = element_text(size = 20, face = "bold"),
        plot.subtitle = element_text(size = 14),
        axis.text.x = element_text(angle = 45, hjust = 1))

# Display the plot
print(plot_category_sales)
```

```{r}
# Create the ggplot2 graph to show the relationship between temperature and clothing sales
plot_clothing_sales_temp <- ggplot(clothing_sales_monthly, aes(x = Average_Temperature, y = Average_Sales)) +
  geom_point(color = 'blue') +
  geom_smooth(method = 'lm', color = 'red', se = FALSE) +
  labs(title = "Relationship Between Temperature and Clothing Sales",
       subtitle = "The chart shows average clothing sales per month against average temperature",
       x = "Average Temperature (°F)",
       y = "Average Clothing Sales (in $)") +
  theme_minimal() +
  theme(plot.title = element_text(size = 20, face = "bold"),
        plot.subtitle = element_text(size = 14))

# Display the plot
print(plot_clothing_sales_temp)
```

## Conclusion

Seasonal Trends: 
Based on the first graph of Average Sales per month for the Walmart and Retail Sales datasets, it is obvious that there is a huge spike around the holiday season (starting around November to December). This makes sense as there are a variety of advertising campaigns during this time period and it is generally a time where people shop around and buy presents for themselves or their families. Retailers are aware of these trends and focus their campaigns around this time period. This trend is confirmed through both datasets which is why I wanted to use two datasets in order to confirm this trend as the Walmart dataset is significantly older. This means that these seasonal trends have been prevelant for at least a decade which creates confidence in this trend in retail sales.

Average sales per month per category: 
The second graph reveals an interesting trend: while there is a significant increase in sales for beauty products during the holiday season, electronic sales experience a sharp drop-off. Beauty products, such as cosmetics and skincare items, are popular gift choices during the holidays. They are often perceived as thoughtful and personal gifts, leading to increased sales in this category.
There are probably a lot of promotions in this area as well. Meanwhile, electronic products see a sharp decline probably because a significant amount of people buy technology during Black Friday and Cyber Monday which lie in late November, causing the dropoff in December where people don't feel the need to do this anymore. 

Temperature impact on clothing sales: 
I extrapolated temperature data from the Walmart dataset to the months that corresponded to the retail sales dataset in order to see the potential impact on clothing sales. Because clothing sales in the second graph were the least impacted by the time of the year, I wanted to see if the temperature was instead a factor in the sale of the clothes. I plotted some points in order to see if there was a trend, and I found that there was a slight correlation between lower temperatures and higher sales in clothing. However, it is not something that can be a direct link as more data would be required and the way that I got my temperatures for this time period was not entirely accurate. However, it does make sense that in lower temperatures, people spend more money on warmer clothing while in warmer temperatures, special kinds of clothing as less required or necessary. 

Conclusion: 
In this research, I examined how seasonal changes and temperature variations impact retail sales across different product categories. This analysis revealed a consistent spike in sales during the holiday season (November to December) across both the synthetic retail sales data and historical Walmart sales data. Beauty products saw significant increases due to their popularity as gifts and promotional efforts, while electronics experienced a drop-off after major sales events like Black Friday and Cyber Monday. Additionally, I found a slight correlation between lower temperatures and higher clothing sales, suggesting that colder weather drives demand for warmer clothing. These findings highlight the importance for retailers to adjust marketing strategies, inventory levels, and promotions based on seasonal trends and temperature variations to optimize sales.

